---
title: "POPLS-DA Computational Workflow for vertical multi-omics data"
output:
  html_document:
    toc: yes
    df_print: paged
#  html_notebook:
#    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

# POPLS-DA Computational Workflow {.tabset .tabset-fade .tabset-pills}

## Background 

These packages are needed for all functionalities

```{r Libraries}
## CRAN or BioConductor packages
library("BiocManager")
library("devtools")
library("OmicsPLS")
library("mixOmics")
library("preprocessCore")
library("edgeR")
library("ropls")
library("org.Hs.eg.db")
library("GO.db")
library("tidyverse")
library("magrittr")
library("ggrepel")
library("gridExtra")
library("STRINGdb")
library("RColorBrewer")
library("rvest")
library("cluster")
library("plotly")
library("kableExtra")
#library("dbparser")

## GitHub packages
library(PO2PLS) # devtools::install_github("selbouhaddani/PO2PLS")
library(disgenet2r) # devtools::install_bitbucket("ibi_group/disgenet2r")
library("dbdataset") # devtools::install_github("interstellar-Consultation-Services/dbdataset")

# This custom goana function is needed 
source("goana_default.R")

## The POPLS-DA Code
source("https://raw.githubusercontent.com/selbouhaddani/POPLSDA/main/POPLS2.R")

```

## Pre-processing multi-omics

First, simulate some artificial multi-omics data. 

```{r Simulate data}
########## Pre-processing multi-omics data #####
## Define the datasets
## Note, this is designed for data on different samples
# Dataset in format all Omics Samples x Genes
N_samples <- 50
p_genes <- 2000
Data_MultiOmics <- matrix(rnorm(N_samples*p_genes), nrow=N_samples, ncol=p_genes)
rownames(Data_MultiOmics) <- 1:nrow(Data_MultiOmics)
# Add colnames of the data are the gene names
colnames(Data_MultiOmics) <- org.Hs.eg.db %>% 
  AnnotationDbi::select(keys=keys(org.Hs.eg.db)[1:p_genes], 
                        columns = "SYMBOL") %>% pull("SYMBOL")
# Define the outcome/experimental groups for all samples
Data_outcome <- rbinom(nrow(Data_MultiOmics), 1, prob = 0.5)
# Define the type of omics data for each sample
Data_OmicsLevels <- factor(sort(rep(1:2, length.out = nrow(Data_MultiOmics))), 
                           labels = c("Omics_1","Omics_2"))
Data_OmicsLvls <- as.numeric(Data_OmicsLevels)
```

Pre-process these data.

```{r Pre-process}
## Scale
for(ii in 1:length(levels(Data_OmicsLevels))){
  Data_MultiOmics[which(Data_OmicsLvls == ii),] %<>% scale
}

## Reference association profile
ref_test <- 
  sapply(1:ncol(Data_MultiOmics), 
     function(ii) {
       coef(summary(lm(
         Data_MultiOmics[which(Data_OmicsLvls == 1),ii] ~ 
           Data_outcome[which(Data_OmicsLvls==1)])))[2,3] %>% sign
     })

## Align for other omics their association profile
for(lvl_i in 2:length(levels(Data_OmicsLevels))){
  lvl_test <- 
    sapply(1:ncol(Data_MultiOmics), 
           function(ii) {
             coef(summary(lm(
               Data_MultiOmics[which(Data_OmicsLvls == lvl_i),ii] ~ 
                 Data_outcome[which(Data_OmicsLvls == lvl_i)])))[2,3] %>% sign
           })
  Data_MultiOmics[Data_OmicsLvls==lvl_i,] %<>% 
    sweep(2, ref_test*lvl_test,`*`)
}

```


## POPLS-DA data integration

First, choose the number of components (e.g. via a scree-plot).
```{r Choose nr components}
########## POPLS-DA integrative model  ##### 
# Scree plot and Gap Statistic
plot(svd(Data_MultiOmics,nu=0,nv=0)$d^2 / sum(svd(Data_MultiOmics,nu=0,nv=0)$d^2))
# Initial fit for object structure
fit_splsda <- splsda(X = Data_MultiOmics, Y = Data_outcome, ncomp = 2)
#Prepare POPLS-DA as a list with individual omics datasets
Data_POPLS <- 
  lapply(unique(Data_OmicsLvls),
         function(ii) list(X=Data_MultiOmics[Data_OmicsLvls==ii,], 
                           Y=Data_outcome[Data_OmicsLvls==ii]))
p_genes #nr of genes
K <- length(unique(Data_OmicsLvls)) #nr of omics levels
r <- 2 #nr of joint components
rx <- 1 #nr of specific components 

```

Prepare and fit the POPLS-DA model to the data.

```{r Fit POPLS-DA}
# Generate inital parameters, partly based on initial fit
prm.tst <- with(generate_params(ncol(Data_MultiOmics), 10, r, K*rx, 0, type = "r"), list(
  W = fit_splsda$loadings$X,
  P = lapply(1:K, function(K_ii) matrix(rnorm(ncol(Data_MultiOmics)*rx),ncol=rx)),
  beta = diag(B),
  SigT = SigT,
  SigU = SigTo[1:rx, 1:rx],
  sig2e = as.list(rep(sig2E, length(unique(Data_OmicsLvls)))),
  sig2eps = as.list(rep(SigH[1], length(unique(Data_OmicsLvls))))#,
  #Pen = list(lambda = 0, Target_W = fit_splsda$loadings$X)
))
prm.tst$W %<>% as.matrix %>% multiply_by_matrix(sqrt(prm.tst$SigT))
prm.tst$SigT <- diag(1, r)
prm.cur <- prm.tst#[1:8]

# Fit POPLS-DA
fit_poplsda <- POPLS.EM(Data_POPLS, prm.cur, nsteps = 1e3, atol=1e-7)

```

Plot and inspect the results in terms of top genes and prediction performance. 

```{r Visualize results}
# Rank genes according to effect size
topGenes <- (fit_poplsda$par$W%*%fit_poplsda$par$beta)[,1] %>% abs %>% sort(decreasing=T) %>% tibble(weight=.,SYMBOL=names(.))

## Output of fit
# Plot of samples per component
tibble(T1 = orth(Data_MultiOmics %*% fit_poplsda$param$W[,1]),
       T2 = orth(Data_MultiOmics %*% fit_poplsda$param$W[,2]),
       Group = factor(Data_outcome,labels = c("Case","Control")),
       Omics = Data_OmicsLevels) %>%
  ggplot(aes(x=T1, y=T2)) +
  geom_point(aes(col=Group, shape = Omics), size=3) +
  scale_color_manual(values = c("#E69314", "#0C7BDC")) +
  stat_ellipse(aes(x=T1, y=T2, col=Group), level = 0.9, size=1.2) +
  theme_bw() + theme(text = element_text(size=16)) +
  xlab("Joint scores 1") + ylab("Joint scores 2")
  
# Explained variances by joint parts
sapply(levels(Data_OmicsLevels), function(ee){
  ii <- which(levels(Data_OmicsLevels) == ee)
  R2_toti <- with(fit_poplsda$param, 
                  tr(crossprod(W))+tr(crossprod(P[[ii]]))+ncol(Data_MultiOmics)*sig2e[[ii]])
  with(fit_poplsda$param, (tr(crossprod(W))+tr(crossprod(P[[ii]])))/R2_toti)
}, USE.NAMES = TRUE)

# Select the number of relevant genes with a scree plot
eff_sze <- with(fit_poplsda$par, W %*% beta)
eff_sze.srt <- orth(eff_sze[order(-eff_sze^2)])
tibble(nrGns = seq_along(eff_sze), VarExpl=eff_sze.srt^2) %>%
  qplot(x=nrGns, y=VarExpl, data = ., geom="point",
        xlab = "Genes", ylab="Individual variance explained")
top_K <- 200

# After selection, use these genes to predict the outcome
Wbeta_subset <- fit_poplsda$par$W %*% fit_poplsda$par$beta
Wbeta_subset[-order(-Wbeta_subset^2)[1:top_K]] = 0
tibble(Predictions = scale(Data_MultiOmics %*% Wbeta_subset),
       Group = factor(Data_outcome,labels = c("Case","Control")),
       Omics = Data_OmicsLevels) %>%
  ggplot(aes(x=Group, y=Predictions)) +
  geom_boxplot() +
  geom_jitter(aes(col=Omics), width = 0.2, height = 0, size = 3) +
  theme_bw() + theme(text = element_text(size=20)) +
  scale_color_manual(values = c("#E69314", "#0C7BDC"))

```


## Direct Neighbor approach

First, find the drug targets of the given drugs.

```{r}
## Find drug targets of given drugs
refDrugs <- c("Dexibuprofen", "Imatinib") # example drugs
refGenes <- sapply(refDrugs, function(e) {
  cat("\n Looking up ",e,"... \n", sep="")
  DBDrug <- dbdataset::drugbank$drugs$general_information %>% filter(name == e)
  if(nrow(DBDrug) > 0){
    DBhtml <- read_html(paste0("https://www.drugbank.ca/drugs/",DBDrug$primary_key))
    DBtext <- DBhtml %>% html_nodes(css='div.col-sm-12.col-lg-7') %>% html_text
    DBgene.loc <- DBtext %>% lapply(function(ee) ee %>% str_locate(c("Uniprot ", "Gene Name")) %>% diag %>% `[`(c(2,1)))
    if(length(DBgene.loc)>0) {
      DBgene <- sapply(1:length(DBgene.loc), function(ii) DBtext[ii] %>% str_sub(DBgene.loc[[ii]][1]+1, DBgene.loc[[ii]][2]-1))
      cat("  \U2714 -- Gene(s) found: ***", DBgene, "***"); return(DBgene)}
    else {cat("  \U274C -- Drug found, but gene NOT FOUND"); return(NA)}
  } else {cat("  \U274C -- DRUG NOT FOUND"); return(NA)}
}) %>% na.omit %>% c
cat("\n")
if(anyNA(refGenes)) refGenes <- refGenes[-which(is.na(refGenes))]

```

Initialize a String-DB connection to find interacting neighbors.

```{r}
# Initialize a string-db connection, needs internet
if(!exists("string_db")) string_db <- STRINGdb$new( version="11.5", species=9606, score_threshold=400, input_directory="")
# Map top Genes to string identifiers, takes some time
dat.mapped <- string_db$map(as.data.frame(topGenes[1:top_K,]), "SYMBOL", removeUnmappedRows = FALSE )
if(anyDuplicated(dat.mapped$SYMBOL) > 0) dat.mapped %<>% filter(!(duplicated(SYMBOL)))
# Calculate all neighbors of the top genes, takes some time
# Rerun when connection time out
tophits <- dat.mapped %>% pull(STRING_id)
tophits.nb <- c(string_db$get_neighbors(tophits), tophits) %>% unique
dat.top <- dat.mapped 

```

Now, calculate the direct neighbors of each gene/drug. 

```{r Find DNs}
## Direct Neighbor approach
# Integrate drug targets into top genes from POPLS-DA
# Identify overlap between top genes (+neighbors) and drug targets
tophits_nball <- lapply(seq_along(refGenes), function(ii) {
  refdrug <- dbdataset::drugbank$drugs$general_information %>% filter(primary_key == names(refGenes[ii])) %>% pull("name")
  print(paste0(ii, "/",length(refGenes),"  ", refdrug))
  refgene <- na.omit(refGenes[[ii]])
  PDE1 <- string_db$mp(refgene)
  PDE1.nb <- string_db$get_neighbors(PDE1)
  print(paste0("Nr of interactors:", length(tophits.nb %>% intersect(PDE1.nb))))
  tophits.pde1 <- tophits %>% intersect(PDE1.nb) %>% c(PDE1)
  return(dat.top %>% filter(STRING_id %in% tophits.pde1) %>% pull(SYMBOL))
})
refGenes_names <- names(refGenes)
names(tophits_nball) <- refGenes_names
# Create a gene x drug table with an indicator (0/1) of a Direct Neighbor
gene_drug_tbl <- sapply(tophits_nball, function(e) dat.top$SYMBOL %in% e) %>% 
  as_tibble %>% `+`(0)
gene_drug_tbl %<>% bind_cols(SYMBOL = dat.top$SYMBOL, .) %>% as_tibble

```

Print the drug by DN (direct neighbors) table, as well as in LaTeX code.

```{r}
# Print the table in LaTeX table format
tmp <- gene_drug_tbl %>% select(-1) %>% apply(2, sum) %>% sort(decreasing = T)
tmp <- data.frame(Rank = 1:length(tmp), Drugs = names(tmp), DNs = tmp)
tmp %>% as_tibble %>% kbl %>% 
  kable_paper("hover") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),fixed_thead = T) %>% 
  scroll_box(width = "80%", height = "500px")
lapply(1:nrow(tmp), function(ii){
  cat(unlist(tmp[ii,]), sep=" & ")
  cat(" \\\\ \n")
}) %>% invisible

```

Print the gene by drug table, as well as in LaTeX code.

```{r}
# Arrange the table sorted by number of DNs per gene
gene_drug_tbl %>% mutate(NrDrugs = rowSums(gene_drug_tbl[,-1])) %>% select(1,NrDrugs) %>% 
  arrange(-NrDrugs) %>% kbl %>% 
  kable_paper("hover") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),fixed_thead = T) %>% 
  scroll_box(width = "80%", height = "500px")

# Arrange and print the targeting drugs per gene in LaTeX format
GenexDrug <- lapply(1:nrow(gene_drug_tbl), function(e){
  tibble(SYMBOL = gene_drug_tbl$SYMBOL[e], 
         DRUGS = str_flatten(colnames(gene_drug_tbl[,-1])[which(gene_drug_tbl[e,-1] == 1)], collapse = "; "))
}) %>% Reduce(f=bind_rows)
apply(GenexDrug, 1, function(e) {cat(e, sep=" & "); cat("\\\\"); cat("\n")})

```


## Integrated Interaction Network

First, map the number of DNs to a color profile.  
```{r}
# Generate colors based on nr of targets
FDAcolr <- colorRampPalette(c("white","green"))
plotCol <- sapply(1:nrow(gene_drug_tbl), function(ii) {
  FDAcol = sum(gene_drug_tbl[ii,-1])
  out <- NA
  if(FDAcol > 0) out = FDAcolr(max(rowSums(gene_drug_tbl[,-1]))+1)[FDAcol+1]
  out
})
dat.top$color <- plotCol
# payload_id <- string_db$post_payload(tophits, colors=dat.top$color[1:top_K])
# string_db$plot_network(tophits, payload_id=payload_id)

tophits.conn <- string_db$get_interactions(tophits)[1:2] %>% 
  Reduce(f = c) %>% unique
#par(mar=c(.1,.1,.1,.1))
#string_db$plot_network(tophits.conn, add_link = F, add_summary = F)
#par(mar = .pardefault$mar)
```

Print output format for use as payload on the String-DB website.

```{r}
######## List for payload ######
sweblist <- cbind(tophits, dat.top$color, "-")[1:top_K,]
sweblist <- cbind(sweblist, 
                  mapIds(org.Hs.eg.db, as.character(dat.top$SYMBOL), 
                         "UNIPROT", "SYMBOL", multiVals="first"))
if(anyNA(sweblist[,1])) sweblist <- sweblist[!is.na(sweblist[,1]),]
sweblist[which(is.na(sweblist[,2])),2] <- "#F7F6F2"
#sweblist[,4] <- paste0("https://www.uniprot.org/uniprot/", sweblist[,4])
sweblist <- cbind(sweblist, "-")
invisible(apply(sweblist,1,function(e) {cat(e,sep='\t'); cat('\n')}))

```



## Functional enrichment of relevant genes

First, the necessary preliminary functions are defined.
```{r}
########## Functional enrichment analyses  ##### 
## Function to map IDs to Symbols 
ID2Symb <- function(e, from = "ENTREZID", to = "SYMBOL"){
  AnnotationDbi::select(org.Hs.eg.db, e, keytype = from, columns = to) %>% 
    pull(to)
}

# This custom goana function is needed 
source("goana_default.R")

```

Perform a GO, KEGG and DisGeNet enrichment analysis and print the gene by term tables.

```{r}
topEntrez <- topGenes[1:top_K, ]$SYMBOL %>% 
  ID2Symb(from = "SYMBOL", to = "ENTREZID")

topEntrez %>% goana.default -> GOtbl

GOtbl$tbl[,-1] %>% as.data.frame %>% topGO %>% kbl %>%
  kable_paper("hover") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),fixed_thead = T) %>%
  scroll_box(width = "100%", height = "500px")

topEntrez %>% kegga -> KGtbl

KGtbl %>% topKEGG %>% kbl %>%
  kable_paper("hover") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),fixed_thead = T) %>%
  scroll_box(width = "100%", height = "500px")

GOtbl.top <- GOtbl$tbl[,-1] %>% as.data.frame %>% topGO
GOtbl.top <- inner_join(GOtbl.top, GOtbl$tbl)
names(GOtbl$gns)[2] <- "GOID"

GOtbl.gns <- GOtbl$gns %>% filter(gene_id %in% (topEntrez))
GOtbl.top2 <- inner_join(GOtbl.top, GOtbl$gns[,1:2])
GOtbl.mrx <- sapply(simplify = F, 
                    GOtbl.top2$GOID %>% unique, 
                    function(e) GOtbl.top2 %>% filter(GOID==e) %>% pull(gene_id) %>% ID2Symb)
gns_all <- GOtbl.mrx %>% unlist %>% unique
GOtbl.mrx2 <- sapply(names(GOtbl.mrx), function(e) {
  gns <- GOtbl.mrx[[e]]
  gns_all %in% gns
})

GOtbl.mrx2 %<>% bind_cols(SYMBOL=gns_all) %>% dplyr::inner_join(topGenes[1:top_K,"SYMBOL"],.) %>%
  mutate(SYMBOL=factor(SYMBOL,levels=SYMBOL,labels=SYMBOL))
GOtbl.mrx2 %<>% bind_cols(Rank = 1:nrow(.), .)

GOtbl.plot <- GOtbl.mrx2 %>% as_tibble %>%
  bind_cols(colr=rep(c("#E69314", "#0C7BDC"),length.out=nrow(GOtbl.mrx2))) %>% pivot_longer(contains("GO:"))
GOtbl.plot %<>% mutate(name=factor(name, name, name))
GOtbl.plot %<>% mutate(colr = ifelse(value, colr, "white"),
                       Term = AnnotationDbi::select(GO.db,
                                                    keys=as.character(name), keytype="GOID", columns="TERM")[[2]],
                       GName = AnnotationDbi::select(org.Hs.eg.db,
                                                     keys=as.character(SYMBOL), keytype="SYMBOL", columns="GENENAME")[[2]])
names(GOtbl.plot) <- c("Rank","GeneSymbol", "colr", "GOid", "value","GOterm","GeneName")
GOtbl.plot$GeneSymbol %<>% fct_rev
p <- GOtbl.plot %>% ggplot(aes(x=GOid, y=GeneSymbol,
                               GENElabel=GeneName,GOlabel=GOterm,Rank=Rank)) +
  geom_point(size=GOtbl.plot$value*3, col=GOtbl.plot$colr) +
  theme_bw() + scale_x_discrete(position = "top") + theme(axis.text.x = element_text(angle=45)) +
  xlab("GO ID") + ylab("Top Genes")
p %>% ggplotly(tooltip = c("Rank","GeneSymbol","GENElabel","GOlabel"), height=3000) %>%
  layout(xaxis = list(side ="top" ) )


## DisGeNet analysis
tophits.disgenet <- 
  disease_enrichment(entities = topGenes$SYMBOL[1:top_K],database = "BEFREE" )
tophits.disgenet@qresult[1:20, c("Description", "FDR", "Ratio",  "BgRatio")]

```

